\gloss{CPS}{see \emph{continuation-passing style}}

\gloss{Coercible}{a typeclass showing two types have the same representation at runtime}

\gloss{Contravariant}{a typeclass for ``backwards''-\ty{Functor}s}

\gloss{Covariant}{another name for \ty{Functor}}

\gloss{FCF}{see \emph{first class family}}

\gloss{GADTs}{generalized algebraic data types. Useful for giving explicit, restrictive types on data constructors. Enabled via \ext{GADTs}}

\gloss{indexed monad}{a monadic structure which carries a piece of static state along with it. Indexed monads allow you to enforce protocols in the type system}

\gloss{MPTC}{a multi-parameter typeclass. Enabled via \ext{MultiParamTypeClasses}}

\gloss{OverloadedLabels}{syntax for converting \kind{Symbol}s into values. Used via the syntax \hs{#mySymbol}, and desugared in terms of the \hs{GHC.Overloadedlabels.fromLabel} function. Enabled via \ext{OverloadedLabels}}

\gloss{RankNTypes}{an extension which enables writing higher-rank types. Enabled
via \ext{RankNTypes}}

\gloss{ST trick}{a technique for scoping the lifetime of a piece of data via an existential variable}

\gloss{ScopedTypeVariables}{a language extension which brings type variables into scope if their type had an explicit \hs{forall} quantifier. For example, the type \ty{forall a. a -> Int} brings \ty{a} into scope, but the type \ty{a -> Int} doesn't. Enabled via \ext{ScopedTypeVariables}}

\gloss{TypeInType}{a language extension which allows you to use types as kinds, and vice versa. Enabled via \ext{TypeInType}}

\gloss{ad-hoc polymorphism}{another name for the overloading we get from typeclasses}

\gloss{algebraic data type}{any type made up of sums, products and exponents types}

\gloss{ambiguous type}{a type is an ambiguous when it is unable to be inferred from a callsite. See \ext{AllowAmbiguousTypes} and use \ext{TypeApplications} to disambiguate them}

\gloss{associated type family}{a type family associated with a typeclass.  Instances of the class must provide an instance of the type family}

\gloss{bifunctor}{a type which is a functor over its last two type parameters}

\gloss{canonical product}{another name for \ty{(,)}}

\gloss{canonical representation}{every type is isomorphic to its canonical
representation---a type defined as a sum of products}

\gloss{canonical sum}{another name for \ty{Either}}

\gloss{canonical unit}{another name for \ty{()}}

\gloss{cardinality}{the number of unique values inhabiting a type. Two types with the same cardinality are always isomorphic}

\gloss{carrier}{informal name for a typeclass whose only purpose is to carry ad-hoc polymorphic implementations for generic methods}

\gloss{closed type family}{a type family with all of its instances provided in its definition. Closed type families are a close analogue of functions at the type-level}

\gloss{constraint synonym}{a technique for turning a type synonym of \kind{Constraint}s into something partially-applicable. Performed by making a new typeclass with a superclass constraint of the synonym, and giving instances of it for free given the superclass constraint. For example, \hs{class c a => Trick a} and \hs{instance c a => Trick a}}

\gloss{constraint synonym}{a technique for constructing constraints that can be
partially applied. Performed by creating a new typeclass with the desired
synonym as a superclass constraint, and then by giving a free instance of this
class for any types that have the constraint}

\gloss{constraint trick}{the transformation of a multiparameter typeclass
instance from \ty{instance Foo Int b} to \ty{instance (a \tyeqSpace Int) => Foo a b}. Useful for improving type inference when working with MPTCs}

\gloss{constraint}{a type of kind \kind{Constraint}. Constraints are the things
that exist to the left of the fat context arrow (\hs{=>}).}

\gloss{continuation-passing style}{the technique of taking (and subsequently
calling) a callback, rather than directly returning a value}

\gloss{contravariant}{a type \ty{T a} is contravariant with respect to \ty{a} if
it can lift a function \ty{a -> b} into a function \ty{T b -> T a}}

\gloss{correct by construction}{using the type system to ensure that incorrect
programs are inexpressible}

\gloss{covariant}{a type \ty{T a} is covariant with respect to \ty{a} if
it can lift a function \ty{a -> b} into a function \ty{T a -> T b}. Another name
for a \ty{Functor}}

\gloss{defunctionalization}{a technique for replacing a family of functions with an opaque symbol, and moving the original logic into an \hs{eval} function.  Used by \emph{First Class Families}}

\gloss{dependent pair}{a type that pairs a singleton with a value indexed by the
singleton}

\gloss{dependent type}{a type which isn't known statically, which depends on
term-level values}

\gloss{endomorphism}{a function of the form \ty{a -> a}}

\gloss{fcf}{a first class (type) family}

\gloss{first class family}{a technique for building reusable, higher-order type
families via defunctionalization}

\gloss{functional dependency}{an additional invariant added to a multiparameter
typeclass declaration saying that some of its type varaibles are entirely
determined by others. Primarily used to improve type inference}

\gloss{higher rank}{another name for a rank-\emph{n} type}

\gloss{higher-kinded type}{a type which is parameterized by something other than
\kind{Type}}

\gloss{indexed monad}{a monad that passes additional, static information between
subsequent binds}

\gloss{instance head}{the part of a typeclass instance that comes after the
context arrow (\hs{=>})}

\gloss{introduction}{another word for constructor}

\gloss{invariant}{a higher-kinded type is said to be invariant in a type parameter if that parameter is in neither positive nor negative position}

\gloss{isomorphism}{an isomorphism is a mapping between two things---primarily
types in this book. If two types are isomorphic, they are identical for all
intents and purposes}

\gloss{kind error}{like a type error, but at the kind-level}

\gloss{kind signature}{a declaration (inferred or specified) of a type's kind}

\gloss{negative position}{a type variable which is contravariant with respect to
its data type}

\gloss{nominal}{a type variable is at role nominal if it is an error to coerce that type into another type}

\gloss{non-injectivity}{a property of type families. Something that is
non-injective does not have an inverse}

\gloss{overloaded labels}{syntax for converting \kind{Symbol}s into values. Used via the syntax \hs{#mySymbol}, and desugared in terms of the \hs{GHC.Overloadedlabels.fromLabel} function. Enabled via \ext{OverloadedLabels}}

\gloss{parametric polymorphism}{the polymorphism that arises from quantified type
variables}

\gloss{phantom}{a type variable is at role phantom if it can safely be coerced
into any other type. Type parameters are called phantom if they aren't used at
the term-level}

\gloss{positive position}{a type variable which is covariant with respect to its
data type}

\gloss{product type}{any type that contains several other types simultaneously}

\gloss{profunctor}{a type \ty{T a b} is a profunctor if it is contravariant in
\ty{a} and covariant with respect to \ty{b}}

\gloss{promoted data constructor}{the type that results from a data constructor
when lifting its type to the kind level. Enabled via \ext{DataKinds}}

\gloss{rank-n type}{a type that delays the instantiantion of its type variables,
allowing polymorphism to be preserved through a function call}

\gloss{rank-n}{a rank-\emph{n} type}

\gloss{rank}{a function's degree of polymorphism}

\gloss{reflexivity}{the property that an object has when it is in relationship with itself. For example, equality is reflexive, because something is always equal to itself}

\gloss{representationally equal}{two types are representationally equal if they
have identical physical layouts in memory}

\gloss{representational}{a type variable is at role representational if it can
be coerced into other type sthat are representationally equal to it}

\gloss{rigid skolem}{a type variable that is both rigid and a skolem}

\gloss{rigid}{a type that was explicitly specified by a programmer. A type that
was not inferred}

\gloss{role system}{the system that ensures role annotations are not violated}

\gloss{role}{a property of a type variable that describes how a data constructor
that owns it is allowed to be coerced}

\gloss{role signature}{the declared roles for a data type's type parameters}

\gloss{sigma type}{another name for dependent pair}

\gloss{singleton}{a type with a single inhabitant. Can be abused to create an
isomorphism between types and terms}

\gloss{skolem}{an existentially quantified variable}

\gloss{strengthen}{the act of using a stricter role than is necessary}

\gloss{structural polymorphism}{a technique for automating boilerplate code}

\gloss{structural recursion}{tackling a problem by dividing it and conquering
the pieces}

\gloss{structured logging}{logging real data types rather than only their string
representations}

\gloss{sum of products}{a possible form that a type can be expressed in}

\gloss{sum type}{a type with multiple data constructors}

\gloss{symmetry}{the property that two objects have in a relationship when it is
bidirectional. For example, equality is symmetric, because if $a = b$ then
$b = a$
}

\gloss{term}{a value of a type. Something that can exist at runtime}

\gloss{tick}{the leading apostrophe in the name of a promoted data constructor}

\gloss{transitivity}{the idea that if
$a \star b$ and
$b \star c$ then
$a \star c$ for any relation
$\star$
}

\gloss{type family dependency}{a technique for adding injectivity to a type
family}

\gloss{value type}{a type with kind \kind{Type}}

\gloss{variance}{the behavior a type has when transforming its type parameters}

