\documentclass[book.tex]{subfiles}
\begin{document}

\chapter{Custom Type Errors} \label{type errors}
\label{TypeError}

\ty{OpenSum} and \ty{OpenProduct} are impressive when used correctly. But the
type errors that come along with their misuse are nothing short of nightmarish
and unhelpful. For example:

\begin{dorepl}{OpenSum}
let foo = inj (Identity True) :: OpenSum Identity '[Bool, String]
prj foo :: Maybe (Identity Int)
\end{dorepl}

As a user, when we do something wrong we're barraged by meaningless errors about
implementation details. As library writers, this breakdown in user experience is
nothing short of a failure in our library. A type-safe library is of no value if
nobody knows how to use it.

Fortunately, GHC provides the ability to construct custom type errors. The
module \ty{GHC.TypeLits} defines the type \ty{TypeError} of kind
\kind{ErrorMessage -> k}. The semantics of \ty{TypeError} is that if GHC is ever
asked to solve one, it emits the given type error instead, and refuse to
compile. Because \ty{TypeError} is polykinded, we can put it anywhere we'd like
at the type-level.

The following four means of constructing \kind{ErrorMessage}s are available to
us.

\newcommand{\tyerror}[3]{\item{\ty{'#1} (of kind \kind{#2}.) #3}}

\begin{itemize}
  \tyerror{Text}{Symbol -> ErrorMessage}{Emits the symbol verbatim. Note that this is \emph{not} \ty{Data.Text.Text}.}
  \tyerror{ShowType}{k -> ErrorMessage}{Prints the name of the given type.}
  \tyerror{(:<>:)}{ErrorMessage -> ErrorMessage -> ErrorMessage}{Concatenate two \kind{ErrorMessage}s side-by-side.}
  \tyerror{(:\$\$:)}{ErrorMessage -> ErrorMessage -> ErrorMessage}{Append one \kind{ErrorMessage} vertically atop another.}
\end{itemize}

\ty{TypeError} is usually used as a constraint in an instance context, or as the
result of a type family. As an illustration, we can provide a more helpful error
message for showing a function. Recall that the usual error message is not
particularly useful:

\begin{dorepl}{OpenProduct}
show id
\end{dorepl}

However, by bringing the following instance into scope:

\snip{Misc}{pragmas}
\snip{Misc}{showFunc}

Attempting to show a function in GHCi now offers a solution to what might be
wrong:

\begin{dorepl}{Misc}
show id
\end{dorepl}

When evaluating \hs{show} \hs{id}, Haskell matches the instance head of \ty{Show
(a -> b)}, and then attempts to solve its context. Recall that whenever GHC sees
a \ty{TypeError}, it fails with the given message. We can use this principle to
emit a friendlier type error when using \hs{prj} incorrectly.

\snip{OpenSum}{FriendlyFindElem}

Notice that \ty{FriendlyFindElem} is defined as a \emph{type family}, rather
than a \emph{type synonym} as FCFs usually are. This is to delay the expansion
of the type error so GHC doesn't emit the error immediately. We now attempt to
find \ty{t} in \ty{ts}, and use \ty{FromMaybe} to emit a type error in the case
that we didn't find it.

Rewriting \hs{prj} to use \ty{KnownNat (FriendlyFindElem t ts f ts)} instead of
\ty{Member t ts} is enough to fix our error messages.

\begin{dorepl}{OpenSum}
let foo = inj (Identity True) :: OpenSum Identity '[Bool, String]
friendlyPrj foo :: Maybe (Identity Int)
\end{dorepl}

Let's return to the example of \hs{insert} for \ty{OpenProduct}. Recall the
\ty{UniqueKey key ts \tyeqSpace 'True} constraint we added to prevent duplicate keys.

\snipRename{OpenProduct}{oldInsert}{insert}

This is another good place to add a custom type error; it's likely to happen,
and the default one GHC will emit is unhelpful at best and horrendous at worst.
Because \ty{UniqueKey} is already a type family that can't get stuck (ie. is
total), we can write another type family that will conditionally produce the
error.

\snip{OpenProduct}{RequireUniqueKey}

\ty{RequireUniqueKey} is intended to be called as \ty{RequireUniqueKey
(UniqueKey key ts) key t ts}. The \kind{Bool} at \ann{1} is the result of
calling \ty{UniqueKey}, and it is pattern matched on. At \ann{2}, if it's
\ty{'True}, \ty{RequireUniqueKey} emits the unit constraint \ty{()}
\footnote{This requires \extension{ConstraintKinds}.}. As a \kind{Constraint},
\ty{()} is trivially satisfied.

Notice that at \ann{3} we helpfully suggest a solution. This is good form in
any libraries you write. Your users will thank you for it.

We can now rewrite \hs{insert} with our new constraint.

\snip{OpenProduct}{insert}

\begin{exercise}
Add helpful type errors to \ty{OpenProduct}'s \hs{update} and
\hs{delete} functions.
\end{exercise}
\begin{solution}
\unspacedSnip{OpenProduct}{FriendlyFindElem}
\snipRename{OpenProduct}{friendlyUpdate}{update}
\snipRename{OpenProduct}{friendlyDelete}{delete}

These functions could be cleaned up a little by moving the \ty{FriendlyFindElem}
  constraint to \ty{findElem}, which would remove the need for both constraints.
\end{solution}

\begin{exercise}
Write a closed type family of kind \kind{[k] -> ErrorMessage} that
pretty prints a list. Use it to improve the error message from
\ty{FriendlyFindElem}.
\end{exercise}
\begin{solution}
\unspacedSnip{OpenProduct}{ShowList}
\snip{OpenProduct}{FriendlyFindElem2}
\end{solution}

\begin{exercise}
See what happens when you directly add a \ty{TypeError} to the context
of a function (eg. \hs{foo :: TypeError ... => a}). What happens? Do you know
why?
\end{exercise}
\begin{solution}
GHC will throw the error message immediately upon attempting to compile the
  module.

The reason why is because the compiler will attempt to discharge any extraneous
  constraints (for example, \ty{Show Int} is always in scope, and so it can
  automatically be discharged.) This machinery causes the type error to be seen,
  and thus thrown.
\end{solution}

\end{document}
