\backmatter

\iffalse
\chapterNoTOC{Glossary}
\input{glossary.tex}
\fi
\printglossaries

\chapterNoTOC{Solutions}
\loadSolutions

\begin{thebibliography}{9}
\addcontentsline{toc}{chapter}{Bibliography}

\bibitem{inspection-testing}{
    Joachim Breitner. \\ \website{https://github.com/nomeata/inspection-testing}}

\bibitem{singletons}{
    Richard Eisenberg. \\ \website{https://github.com/goldfirere/singletons}}

\bibitem{contravariant}{
    Ed Kmett. \\ \website{https://github.com/ekmett/contravariant}}

\bibitem{indexed}{
    Ed Kmett. \\ \website{https://github.com/reinerp/indexed}}

\bibitem{invariant}{
    Ed Kmett. \\ \website{https://github.com/ekmett/invariant}}

\bibitem{kan-extensions}{
    Ed Kmett. \\ \website{https://github.com/ekmett/kan-extensions}}

\bibitem{do-notation}{
    Sandy Maguire. \\ \website{https://github.com/isovector/do-notation}}

\bibitem{one-hole}{
    Conor McBride. ``The Derivative of a Regular Type is its Type of One-Hole
    Contexts.'' \\ \website{http://strictlypositive.org/diff.pdf}}

  \bibitem{aeson}{
      Brian O'Sullivan. \\ \website{https://github.com/bos/aeson}}

  \bibitem{fcf}{
      \lyxia. \\ \website{https://github.com/Lysxia/first-class-families}}

  \bibitem{servant}{
      Zalora South East Asia Pte Ltd. \\ \website{https://github.com/haskell-servant/servant}}

\end{thebibliography}



\chapterNoTOC{About the Author}

Sandy might best be described somewhere between independent researcher and
voluntarily-unemployed bum. At the ripe old age of 27, he decided to quit his
highly-lucrative engineering job and to focus more on living than on grinding
for the man. It's what you might call a work in progress.

In his now-copious spare time, he builds clocks, writes software libraries,
designs video games, and plays lots of music. He maintains a blog about
programming in Haskell at \website{reasonablypolymorphic.com}.

Some things he's done in past lives include being an early engineer at
Formation, building blazingly-fast compilers at Google, and accidentally making
Facebook a lot of money. Despite being retired, he might be interested in
helping you with any of these things too, so get in touch.

He holds a Bachelors Degree in Honors Software Engineering from the University
of Waterloo, which is a fantastic school that he'd recommend you too attend.

Originally he comes from very northern Canada, but has since lived in eleven
cities throughout four different countries. Currently he lives in Ottawa,
Canada.

Want to talk? His email is \email{sandy@sandymaguire.me}.

